### 2021-10-06 paper.Rmd
some stuff

    Anticlustering refers to a collection of methods that partition data sets into disjunct groups while addressing two objectives: (a) high between-group similarity and (b) high within-group heterogeneity. For some anticlustering formulations, these two objectives coincide: high between-group similarity is achieved when within-group heterogeneity is maximized. Other anticlustering methods focus on one of the two aspects. The present paper focuces on maximizing between-group similarity by introducing a novel bicriterion objective for anticlustering. *K-plus anticlustering* is an extension of k-means anticlustering, which minimizes the difference between the mean of numeric attributes across clusters. The k-plus objective extends k-means by adding a term that aims to optimize similarity with regard to the variance of the attributes data across clusters. It is shown that the combined bicriterion objective can be reduced to an augmentation of the input data, leaving the original k-means criterion unchanged. A computer simulation and practical examples on real data show that k-plus anticlustering usually achieves high similarity both with regard to mean attribute values as well as the variance, with no noticable sacrifice to either criterion. The k-plus extension is therefore preferred over classical k-means anticlustering for optimizing between-group similarity. K-plus anticlustering is easily accessible to researchers and practitioners: the method has been implemented as part of the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering is useful in many tasks surrounding research work in psychology. For example, anticlustering can be used to split a test into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually suboptimal. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020]. For a general overview of anticlustering approaches and applications in Psychology, see @brusco2019 and @papenberg2020.

Anticlustering methods can be distinguished by the input data (pairwise dissimilarity vs. attributes values) and the objective function that is used to optimize group between-group similarity or within-group heterogeneity. The most prominent criterion on the basis of pairwise dissimilarity ratings is the *diversity*, i.e., the total sum of dissimilarities between any elements within the same group. Maximizing the diversity simultaneously optimizes within-group heterogeneity and between-group similarity. Because the diversity is the reversal of the clustering editing objective [@zahn1964; @shamir2004vf], @papenberg2020 used the term *anticluster editing* to refer to anticlustering approaches that optimize the diversity. Another important criterion based on pairwise dissimilarities is the *dispersion*, i.e., the minimum dissimilarity between any two elements within the same group [@fernandez2013; @brusco2019]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements as dissimilar from each other as possible.

Oftentimes, the data available to researchers does not represent pairwise similarity, but numeric, categorical or binary attribute values (or *features*; @dry2009). That is, the data is a two-way two-mode [@arabie1992] table where rows represent the entities of study, and columns represent variables describing the elements. For example, numeric ratings may describe the attributes of word stimuli for psycholinguistic experiments. Such data may include norms for imagery, concreteness and orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a facial feature [@tversky1977]. In the context of anticlustering, one possible approach for handling two-way two-mode data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across a set of numeric attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into similar groups.

A different approach to anticlustering directly works with the two-way two-mode feature data by maximizing the k-means criterion. K-means is probably the best-known clustering method. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful anticlustering tool by Späth [-@spath1986] and Valev [-@valev1998; -@valev1983]. In k-means clustering, the sum of the squared Euclidean distances between elements and the center of the cluster they are assigned to is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. For the anticlustering application, the k-means criterion is maximized instead, thereby achieving similarity between groups. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. Hence, groups are similar with regard to the mean of the distribution of the numeric attributes wheras other parameters---such as the variance---are ignored. When requiring that groups are overall similar to each other, this focus may be misguided as similar means can be obtained even when the underlying distributions are different [e.g., @anscombe1973]. @papenberg2020, showed that maximizing the diversity (based on the Euclidean distance) is better suited to minimize difference with regard to both the mean and the variance of the data; it was shown that even an entirely random split usually resulted in more similar variances because k-means solely focuses on the mean similarity.

To optimize overall group similarity---and hence to overcome the limitations of k-means anticlustering---I present k-plus anticlustering in this paper. The k-plus criterion is an extension of the k-means objective that includes a term aiming at optimizing similarity with regard to the variance of data across groups. Interestingly, this bicriterion reformulation can be understood as an augmentation of the original data matrix, leaving the original k-means objective unchanged. I show that the k-plus criterion outperforms other anticlustering criteria when aiming at minimizing differences with regard to the mean and the variance of the different groups. In most cases, both objectives can be optimized satisfactorily at the same time without any loss on either. Moreover, in two pratical examples I illustrate how readers can easily apply k-plus anticlustering using the free and open source software `anticlust`, which is an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

It is unlikely that an exclusive optimization of $SSE_{Var}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $SSE_{Var}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall cluster similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and then combined into a single criterion through a weighted sum:

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of 
equalizing means and variances, respectively. Often, they are chosen such that 
$w_1 + w_2 = 1$ [@brusco2019].

Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. When defining $z_i = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, x_{iP})$, i.e., the concatenation of the $x_{ij}$ and $y_{ij}$ values, the bicriterion $\mathit{SSE_{kplus}}$ is then computed as

@@

This formula again has the same form as $\mathit{SSE}$ (and $SSE_{Var}$), but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending these to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data matrix. Hence, k-plus anticlustering reduces to an augmentation of the data matrix, leaving the original k-means criterion unchanged.

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Since the computation of the data matrix $\mathbf{Y}_{N \times P}$ involves squaring, the squared distances $y_{ij}$ have much larger values, on average, than the raw values $x_{ij}$ (also see Figure 1). Therefore, the $SSE_{Var}$ criterion naturally receives greater weight when computing $\mathit{SSE_{kplus}}$ than the standard $\mathit{SSE}$ criterion. This imbalance is most likely unwarranted because optimizing similarity with regard to variances is not more important than optimizing similarity with regard to means. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

@@

When optimizing a bicriterion objective such as $SSE_{kplus}$, the ideal outcome would be to find a partition where both the means and the variances are very similar between groups, i.e., a which partition where both the standard $\mathit{SSE}$ and $SSE_{Var}$ are close to the global optimum value. Generally, research on multi-objective optimization has shown that cases are rare where a single solution satisfies all criteria to an optimal degree [@ferligoj1992direct]. Usually, a tradeoff is necessary. For anticlustering, a motivating example illustrates that aiming to simultaneously optimize similarity of means and variances is not only desirable, but may actually be feasible.

@@

For this motivating example, I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $SSE_{Var}$ across partitions. Note that for both criteria, a larger value is better, i.e., indicates higher similarity with regard to means and variances, respectively. Therefore, to obtain overall between-group similarity, a partition should be chosen that is depicted in the upper right corner of the scatterplot. However, optimal k-means anticlustering would create two groups that are not particularly similar with regard to their variance. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $SSE_{Var}$. Similarly, the partition that maximizes $SSE_{Var}$ (the triangular symbol in Figure 1) is not well-suited to optimize SSE at the same time. The diamond symbol represents the partition that maximizes (the unweighted) $SSE_{kplus}$, i.e., the sum of $\mathit{SSE}$ and $SSE_{Var}$. This partition is located in the right upper corner and yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $SSE_{Var}$.

>>>>

    Anticlustering refers to a collection of methods that partition data sets into disjunct groups while aiming for two objectives: (a) high between-group similarity and (b) high within-group heterogeneity. For some anticlustering formulations, these two objectives coincide, i.e., high between-group similarity is achieved when within-group heterogeneity is maximized. Other anticlustering criteria focus on one of the two aspects. The present paper focuces on maximizing between-group similarity by introducing a novel bicriterion objective for anticlustering. *K-plus anticlustering* is an extension of k-means anticlustering, which minimizes the difference between the mean of numeric attributes across clusters. The k-plus objective extends k-means by adding a term that aims to optimize similarity with regard to the variance of the attributes data across clusters. It is shown that the combined bicriterion objective can be reduced to an augmentation of the input data, leaving the original k-means criterion unchanged. A computer simulation and practical examples on real data show that k-plus anticlustering usually achieves high similarity both with regard to mean attribute values as well as the variance, with no noticable sacrifice to either criterion. The k-plus extension is therefore preferred over classical k-means anticlustering for optimizing between-group similarity. K-plus anticlustering is easily accessible to researchers and practitioners: the method has been implemented as part of the open source R package `anticlust` and is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering is useful in many tasks surrounding research work in psychology. For example, anticlustering can be used to split a test into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to different---but parallel---experimental conditions [@lahl2006]. For a more comprehensive overview of anticlustering applications in Psychology, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods can be distinguished by the input data (pairwise dissimilarity vs. attributes values) and the objective function that is used to optimize group between-group similarity or within-group heterogeneity. Some anticlustering objectives quantify group similarity on the basis of pairwise dissimilarity rating such as the Euclidean distance. The most prominent criterion on the basis of pairwise dissimilarity ratings is the *diversity*, i.e., the total sum of dissimilarities between any elements within the same group. Maximizing the diversity simultaneously optimizes within-group heterogeneity and between-group similarity. Because the diversity is the reversal of the clustering editing objective [@zahn1964; @shamir2004vf], @papenberg2020 used the term *anticluster editing* to refer to anticlustering approaches that optimize the diversity. Another important criterion based on pairwise dissimilarities is the *dispersion*, i.e., the minimum dissimilarity between any two elements within the same group [@fernandez2013; @brusco2019]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible.

Oftentimes, researchers work with attribute values [or *features*; @dry2009] instead of dissimilarity ratings. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing].  For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into groups. A different approach to anticlustering directly works with the attribute data by maximizing the k-means criterion. K-means is probably the best-known clustering method. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful anticlustering tool by Späth [-@spath1986] and Valev [-@valev1998; -@valev1983]. In k-means clustering, the sum of the squared Euclidean distances between elements and the center of the cluster they are assigned to is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. For the anticlustering application, the k-means criterion is maximized instead, thereby achieving similarity between groups. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. Hence, groups are similar with regard to the mean of the distribution of the numeric attributes wheras other parameters---such as the variance---are ignored. When requiring that groups are overall similar to each other, this focus may be misguided as similar means can be obtained even when the underlying distributions are different [e.g., @anscombe1973]. @papenberg2020, showed that maximizing the diversity (based on the Euclidean distance) is better suited to minimize difference with regard to both the mean and the variance of the data; it was shown that even an entirely random split usually resulted in more similar variances because k-means solely focuses on the mean similarity.

To optimize overall group similarity---and hence to overcome the limitations of k-means anticlustering---I present k-plus anticlustering in this paper. The k-plus criterion is an extension of the k-means objective that includes a term that quantifies similarity with regard to the variance of data across groups. Interestingly, this bicriterion reformulation can be understood as an augmentation of the original data matrix, leaving the original k-means objective unchanged. I show that the k-plus criterion outperforms other anticlustering criteria when the goal is to minimize differences with regard to the mean and the variance of groups. In most cases, both objectives can be optimized satisfactorily at the same time without any loss on either. Moreover, in two pratical examples I illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

It is unlikely that an exclusive optimization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall cluster similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and then combined into a single criterion through a weighted sum:

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. When defining $z_i = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, x_{iP})$, i.e., the concatenation of the $x_{ij}$ and $y_{ij}$ values, the bicriterion $\mathit{SSE_{kplus}}$ is then computed as

@@

This formula again has the same form as $\mathit{SSE}$ (and $\mathit{SSE_{Var}}$), but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending these to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data matrix. Hence, k-plus anticlustering reduces to an augmentation of the data matrix, leaving the original k-means criterion unchanged.

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Since the computation of the data matrix $\mathbf{Y}_{N \times P}$ involves squaring, the squared distances $y_{ij}$ have much larger values, on average, than the raw values $x_{ij}$ (also see Figure 1). Therefore, the $\mathit{SSE_{Var}}$ criterion naturally receives greater weight when computing $\mathit{SSE_{kplus}}$ than the standard $\mathit{SSE}$ criterion. This imbalance is most likely unwarranted because optimizing similarity with regard to variances is not more important than optimizing similarity with regard to means. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

@@

When optimizing a bicriterion objective such as $SSE_{kplus}$, the ideal outcome would be to find a partition where both the means and the variances are very similar between groups, i.e., a which partition where both the standard $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are close to the global optimum value. Generally, research on multi-objective optimization has shown that cases are rare where a single solution satisfies all criteria to an optimal degree [@ferligoj1992direct]. Usually, a tradeoff is necessary. However, a motivating example illustrates that aiming to simultaneously optimize similarity of means and variances is not only desirable, but may actually be feasible.

@@

For this example I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. Note that for both criteria, a larger value is better, i.e., indicates higher similarity with regard to means and variances, respectively. Therefore, to obtain overall between-group similarity, a partition should be chosen that is depicted in the upper right corner of the scatterplot. However, optimal k-means anticlustering would create two groups that are not particularly similar with regard to their variance. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (the triangular symbol in Figure 1) is not well-suited to optimize SSE at the same time. The diamond symbol represents the partition that maximizes (the unweighted) $SSE_{kplus}$, i.e., the sum of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$. This partition is located in the right upper corner and yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$.
